---
title: "Oak Data Exploration"
format: html
editor: visual
---

## Oak exploratory data analysis and cleaning

```{r}
library(tidyverse)

library(readxl)
library(ggplot2)
#install.packages("plyr")
library(plyr)

#import the correct sheet/excel
Survival_HeightApril26Blocks7_6Clean <- read_excel("Oak Project/Data/Survival_HeightApril26Blocks7&6Clean.xlsx", 
                                                   sheet = "Input1", skip=1)
View(Survival_HeightApril26Blocks7_6Clean)

names(Survival_HeightApril26Blocks7_6Clean)

#import the rest of the data
oak_data<- list.files(path= "Oak Project/Data",
                      pattern= "6Clean.xlsx",
                      full.names=TRUE) |> 
  lapply(read_excel,
         na= "NA", sheet= "Input1", skip=1) |> 
  bind_rows()

oak_data

```

Rename rows

```{r}
Survival_HeightApril26Blocks7_6Clean<- Survival_HeightApril26Blocks7_6Clean |> 
  rename(Date= "MM/DD/YYYY",
                                              Page_Number = "Page #",
                                              Recorder=  "Name",
                                              Block= "#...4",
                                              Tag_Num=  "#...5",
                                              Oak_type=  "NR, S, or A",
                                              Treatment= "AC, OA, NP, DT, or DL",
                                              OA_microclimate= "TH, TI, CN",
                                              Seedling_Ht_cm= "(cm)",
                                            Num_leaves=  "#...10",
                                              Num_dead_leaves= "#...11",
                                              Leaf_greenness=  "%",
                                              Num_germinated= "#...13",
                                              RCD=  "(mm)",
                                              Notes= "as written on data sheet")
                                              
  
head(Survival_HeightApril26Blocks7_6Clean)

```

### Summarize data

Change numeric columns to numeric

```{r}
head(Survival_HeightApril26Blocks7_6Clean)

Survival_HeightApril26Blocks7_6Clean<-  Survival_HeightApril26Blocks7_6Clean |> 
  mutate(Seedling_Ht_cm = as.numeric(Seedling_Ht_cm)) |> 
  mutate(Num_leaves = as.numeric(Num_leaves)) |> 
  mutate(Num_dead_leaves = as.numeric(Num_dead_leaves))|> 
  mutate(Leaf_greenness = as.numeric(Leaf_greenness)) |> 
  mutate(Num_germinated= as.numeric(Num_germinated)) |> 
  mutate(RCD=as.numeric(RCD))
```

Summarize amounts of data

```{r}
Survival_HeightApril26Blocks7_6Clean |> 
  group_by(Oak_type) |> count()

Survival_HeightApril26Blocks7_6Clean |> 
  group_by(Treatment) |> count()

Survival_HeightApril26Blocks7_6Clean |> 
  group_by(OA_microclimate) |> count()

#Number of treatments/oak types
Survival_HeightApr_treatmentsum<- Survival_HeightApril26Blocks7_6Clean |> 
  group_by(Oak_type, Treatment) |> count()

head(Survival_HeightApril26Blocks7_6Clean)
```

Generate histograms of numeric data

```{r}
ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = Seedling_Ht_cm) +
  geom_histogram()

ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = Num_leaves) +
  geom_histogram()

ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = Num_dead_leaves) +
  geom_histogram()

ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = Leaf_greenness) +
  geom_histogram()

ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = Num_germinated) +
  geom_histogram()

ggplot(Survival_HeightApril26Blocks7_6Clean) + 
  aes(x = RCD) +
  geom_histogram()
```

Mean, max, min of numeric values

Summarize based on oak type and treatment. What are the main variables of interest?

```{r}

# Create a summary table
Survival_HeightApril26Blocks7_6Clean_summarytbl <- Survival_HeightApril26Blocks7_6Clean |> 
  group_by(Oak_type, Treatment) |> 
  summarize(
    n = n(), 
    min_SeedHt= min(Seedling_Ht_cm,na.rm=TRUE),
    mean_SeedHt = mean(Seedling_Ht_cm, na.rm=TRUE), 
    max_SeedHt= max(Seedling_Ht_cm, na.rm=TRUE),
    SD_SeedHt = sd(Seedling_Ht_cm, na.rm=TRUE)) 

Survival_HeightApril26Blocks7_6Clean_summarytbl
```

Summarize how many seedlings survived out of total planted

```{r}
Survival_HeightApril26Blocks7_6Clean |> 
  summarise(
    seed_count = sum(!is.na(Seedling_Ht_cm)),
    na_count      = sum(is.na(Seedling_Ht_cm))
  )

#29 seedlings survived, 8 did not, 78% survival

29/(8+29)
```

### All data from all sheets

```{r}
library(tidyverse)

#add dplyr to deprecate plyr and call tidyverse function instead
oak_data<- oak_data |> 
  dplyr::rename(Date= "MM/DD/YYYY",
         Page_Number = "Page #",
         Recorder=  "Name",
         Block= "#...4",
         Tag_Num=  "#...5",
         Oak_type=  "NR, S, or A",
         Treatment= "AC, OA, NP, DT, or DL",
         OA_microclimate= "TH, TI, CN",
         Seedling_Ht_cm= "(cm)",
         Num_leaves=  "#...10",
         Num_dead_leaves= "#...11",
         Leaf_greenness=  "%",
         Num_germinated= "#...13",
         RCD=  "(mm)",
         Notes= "as written on data sheet")

detach("package:plyr", unload = TRUE)

 unique(oak_data$Block)
 
 #missing block 3
```

Convert to numeric

```{r}

oak_data<-  oak_data |> 
  mutate(Seedling_Ht_cm = as.numeric(Seedling_Ht_cm)) |> 
  mutate(Num_leaves = as.numeric(Num_leaves)) |> 
  mutate(Num_dead_leaves = as.numeric(Num_dead_leaves))|> 
  mutate(Leaf_greenness = as.numeric(Leaf_greenness)) |> 
  mutate(Num_germinated= as.numeric(Num_germinated)) |> 
  mutate(RCD=as.numeric(RCD))
```

Number germinated column (should just be one seedling per row)

Whether tag number aligns with \# germinated, tag #= plot #, needs to have plant \# instead (split tag #)

```{r}

#take one column per acorn num and tag num
oak_data_clean<- oak_data |>    separate(Tag_Num, into = c("Tag_Num", "Acorn_Num"), sep= "A") |> mutate(Tag_Num= as.numeric(Tag_Num),
             Acorn_Num= as.numeric(Acorn_Num)) |> 
  #mutate(Seedling_Ht_cm= case_when(Seedling_Ht_cm =NA ~0)) |> 
  mutate(Dead= case_when(Seedling_Ht_cm == 0 ~ "Dead",
                         Seedling_Ht_cm >0 ~ "Alive"))


```

```{r}
oak_data_clean |>  
  select(Tag_Num, Acorn_Num,  Block, Treatment, OA_microclimate) |>  
  distinct() |> 
  arrange(Tag_Num)

#test to see if we can make empty rows for non-germinated acorns
Acorn_ID_Table<- oak_data_clean |> 
  select(Tag_Num, Block, Treatment, OA_microclimate, Date, Seedling_Ht_cm) |> 
  distinct() |> 
  arrange(Tag_Num) |> 
  mutate(Acorn1=1, Acorn2=2, Acorn3=3) |> 
  pivot_longer(cols= c(Acorn1, Acorn2, Acorn3), names_to= "whatever", "values_to"= "Acorn_Num") |> 
  select(-whatever)

oak_data_clean2<- oak_data_clean |> select(-Acorn_Num) |> full_join(Acorn_ID_Table) |> 
  relocate(Acorn_Num, .after = Tag_Num)
```

Quick sum

```{r}
ggplot(oak_data_clean) + 
  aes(x = Seedling_Ht_cm) +
  geom_histogram()

ggplot(oak_data_clean) + 
  aes(x = Num_leaves) +
  geom_histogram()

ggplot(oak_data_clean) + 
  aes(x = Num_dead_leaves) +
  geom_histogram()

ggplot(oak_data_clean) + 
  aes(x = Leaf_greenness) +
  geom_histogram()

ggplot(oak_data_clean) + 
  aes(x = Num_germinated) +
  geom_histogram()

ggplot(oak_data_clean) + 
  aes(x = RCD) +
  geom_histogram()

# Create a summary table
oak_data_clean_summarytbl <- oak_data_clean |> 
  group_by(Oak_type, Treatment) |> 
  summarize(
    n = n(), 
    min_SeedHt= min(Seedling_Ht_cm,na.rm=TRUE),
    mean_SeedHt = mean(Seedling_Ht_cm, na.rm=TRUE), 
    max_SeedHt= max(Seedling_Ht_cm, na.rm=TRUE),
    SD_SeedHt = sd(Seedling_Ht_cm, na.rm=TRUE)) 
```

Bar graphs for height across treatments

```{r}
ggplot(oak_data_clean_summarytbl) +
  aes(x = Oak_type, y= mean_SeedHt, fill= Treatment) + 
  labs(x = "Oak type", y= "Mean Seedling Height (cm)") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  scale_x_discrete(labels = c("A" = "Acorn", 
                              "NR" = "Natural Recruit", 
                              "S" = "Seedling"))+
  geom_col(color= "#071A0B")
```

Box plot across treatments

```{r}

ggplot(oak_data_clean) +
  aes(x = Oak_type, y= Seedling_Ht_cm, fill= Treatment) + 
  labs(x = "Oak type", y= "Seedling Height (cm)") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  scale_x_discrete(labels = c("A" = "Acorn", 
                              "NR" = "Natural Recruit", 
                              "S" = "Seedling"))+
  geom_boxplot(color= "#071A0B", outlier.colour = "red")
```

Import october data

```{r}

Survival_HeightOct2025Clean <- read_excel("Oak Project/Data/Survival_HeightOct2025Clean.xlsx", 
    sheet = "Input1")
```
