---
title: "GerminationRateOak"
format: html
editor: visual
---

## Import data

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
#install.packages("plyr")
#library(plyr)

#import all data
oak_data<- list.files(path= "Oak Project/Data",
                      pattern= "6Clean.xlsx",
                      full.names=TRUE) |> 
  lapply(read_excel,
         na= "NA", sheet= "Input1", skip=1) |> 
  bind_rows()

oak_data<- oak_data |> 
  dplyr::rename(Date= "MM/DD/YYYY",
         Page_Number = "Page #",
         Recorder=  "Name",
         Block= "#...4",
         Tag_Num=  "#...5",
         Oak_type=  "NR, S, or A",
         Treatment= "AC, OA, NP, DT, or DL",
         OA_microclimate= "TH, TI, CN",
         Seedling_Ht_cm= "(cm)",
         Num_leaves=  "#...10",
         Num_dead_leaves= "#...11",
         Leaf_greenness=  "%",
         Num_germinated= "#...13",
         RCD=  "(mm)",
         Notes= "as written on data sheet")

#oct 2025 data

Survival_HeightOct2025Clean <- read_excel("Oak Project/Data/Survival_HeightOct2025Clean.xlsx", 
    sheet = "Input1")
```

Convert to numeric and separate acorn and tag numbers

```{r}

oak_data<-  oak_data |> 
  mutate(Seedling_Ht_cm = as.numeric(Seedling_Ht_cm)) |> 
  mutate(Num_leaves = as.numeric(Num_leaves)) |> 
  mutate(Num_dead_leaves = as.numeric(Num_dead_leaves))|> 
  mutate(Leaf_greenness = as.numeric(Leaf_greenness)) |> 
  mutate(Num_germinated= as.numeric(Num_germinated)) |> 
  mutate(RCD=as.numeric(RCD))

#convert NA seedling height to 0
oak_data$Seedling_Ht_cm[is.na(oak_data$Seedling_Ht_cm)] <- 0

#take one column per acorn num and tag num
oak_data_clean<- oak_data |>    separate(Tag_Num, into = c("Tag_Num", "Acorn_Num"), sep= "A") |> mutate(Tag_Num= as.numeric(Tag_Num),
             Acorn_Num= as.numeric(Acorn_Num)) |> 
  mutate(Dead= case_when(Seedling_Ht_cm == 0 ~ "Dead",
                         Seedling_Ht_cm >0 ~ "Alive"))
```

**Total number of acorns should be 3 per tag number**

Filter for acorns only

```{r}
acorn_data<- oak_data_clean |> 
  filter(Oak_type== "A")

#filter for the most recent tag observation
acorn_data_recent <- acorn_data |> 
  group_by(Tag_Num)  |> 
  filter(Date == max(Date)) |> 
  ungroup()

```

Find number of germinated acorns per tag

```{r}
#Find number of surviving acorns per tag (count tag number where seedling ht>0)
acorn_data_count <- acorn_data_recent |> 
  group_by(Tag_Num) |> 
  mutate(Num_germinated = sum(Seedling_Ht_cm > 0, na.rm = TRUE)) |> 
  ungroup()

#Calculate germination rate per tag

acorn_data_count<- acorn_data_count |> 
  group_by(Tag_Num) |> 
  mutate(Germination_rate= Num_germinated/3)
```

Calculate germination rate per treatment

```{r}
ggplot(acorn_data_count) +
  aes(x = Treatment, y= Germination_rate, fill= Treatment) + 
  labs(x = "Treatment", y= "Germination rate") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  geom_violin(color= "#071A0B", outlier.colour = "red")
```

Summarize germination count per treatment

```{r}
germination_count<- acorn_data_count |> 
  group_by(Treatment) |> 
  summarize(Num_germinated_treatment= sum(Num_germinated))

#bar plot
ggplot(germination_count) +
  aes(x = Treatment, y= Num_germinated_treatment, fill= Treatment) + 
  labs(x = "Treatment", y= "Number germinated") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  geom_col(color= "#071A0B")

#see if same number of planted acorns per treatment
# number germinated/ actual acorns planted per treatment
```

Average germination rate per treatment

```{r}

germination_avg<- acorn_data_count |> 
  group_by(Treatment) |> 
  summarize(avg_germinated_treatment= mean(Germination_rate)*100)


#bar plot
ggplot(germination_avg) +
  aes(x = Treatment, y= avg_germinated_treatment, fill= Treatment) + 
  labs(x = "Treatment", y= "Average germination rate (%)") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  scale_x_discrete(labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area"))+
    #geom_text(aes(label =avg_germinated_treatment), vjust = -0.5) +
  geom_col(color= "#071A0B")
```

Germination totals attempt #2

See if same number of planted acorns per treatment

```{r}
acorn_data_count |> 
  group_by(Treatment) |> 
  distinct(Tag_Num) |> 
  count()

#different amount of tags (acorns) per treatment

oak_data_clean |> 
  group_by(Treatment) |> 
  distinct(Tag_Num) |> 
  count()
```

Number germinated/ actual acorns planted per treatment

```{r}

#acorn summary table
acorn_data_count2<- acorn_data_count |> 
  select(Date, Block, Tag_Num, Oak_type, Treatment, Num_germinated, Germination_rate) |> 
  distinct()

germination_count_treat<- acorn_data_count2 |> 
  group_by(Treatment) |> 
  summarize(Num_germinated_treatment= sum(Num_germinated))

acorn_treat_totals<- acorn_data_count |> 
  group_by(Treatment) |> 
  distinct(Tag_Num) |> 
  count()

acorn_data_treatment<- acorn_data_count |> 
  left_join(acorn_treat_totals, by= "Treatment")

acorn_data_treatment |> 
  group_by(Treatment) |> 
  summarize(total_acorn= n*3, germinate_treat= Num_germinated) |> 
  mutate(acorn_treatment= Num_germinated)

```

Plot

```{r}
#bar plot
ggplot(germination_avg) +
  aes(x = Treatment, y= avg_germinated_treatment, fill= Treatment) + 
  labs(x = "Treatment", y= "Average germination rate (%)") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  scale_x_discrete(labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area"))+
  #geom_text(aes(label =avg_germinated_treatment), vjust = -0.5) +
  geom_col(color= "#071A0B")
```

### GLIM

Binomial distribution – 1 acorn per row, germ vs not
