---
title: "GerminationRateOak"
format: html
editor: visual
---

## Import data

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
#install.packages("plyr")
#library(plyr)

#import all data
oak_data<- list.files(path= "Oak Project/Data",
                      pattern= "6Clean.xlsx",
                      full.names=TRUE) |> 
  lapply(read_excel,
         na= "NA", sheet= "Input1", skip=1) |> 
  bind_rows()

oak_data<- oak_data |> 
  dplyr::rename(Date= "MM/DD/YYYY",
         Page_Number = "Page #",
         Recorder=  "Name",
         Block= "#...4",
         Tag_Num=  "#...5",
         Oak_type=  "NR, S, or A",
         Treatment= "AC, OA, NP, DT, or DL",
         OA_microclimate= "TH, TI, CN",
         Seedling_Ht_cm= "(cm)",
         Num_leaves=  "#...10",
         Num_dead_leaves= "#...11",
         Leaf_greenness=  "%",
         Num_germinated= "#...13",
         RCD=  "(mm)",
         Notes= "as written on data sheet")

#oct 2025 data

Survival_HeightOct2025Clean <- read_excel("Oak Project/Data/Survival_HeightOct2025Clean.xlsx", 
    sheet = "Input1")
```

Convert to numeric and separate acorn and tag numbers

```{r}

oak_data<-  oak_data |> 
  mutate(Seedling_Ht_cm = as.numeric(Seedling_Ht_cm)) |> 
  mutate(Num_leaves = as.numeric(Num_leaves)) |> 
  mutate(Num_dead_leaves = as.numeric(Num_dead_leaves))|> 
  mutate(Leaf_greenness = as.numeric(Leaf_greenness)) |> 
  mutate(Num_germinated= as.numeric(Num_germinated)) |> 
  mutate(RCD=as.numeric(RCD))

#convert NA seedling height to 0
oak_data$Seedling_Ht_cm[is.na(oak_data$Seedling_Ht_cm)] <- 0

#take one column per acorn num and tag num
oak_data_clean<- oak_data |>    separate(Tag_Num, into = c("Tag_Num", "Acorn_Num"), sep= "A") |> mutate(Tag_Num= as.numeric(Tag_Num),
             Acorn_Num= as.numeric(Acorn_Num)) |> 
  mutate(Dead= case_when(Seedling_Ht_cm == 0 ~ "Dead",
                         Seedling_Ht_cm >0 ~ "Alive"))
```

**Total number of acorns should be 3 per tag number**

Filter for acorns only

```{r}

oak_data_clean |> 
  group_by(Treatment) |> 
  distinct(Tag_Num) |> 
  count()

acorn_data<- oak_data_clean |> 
  filter(Oak_type== "A")

#filter for the most recent tag observation
acorn_data_recent <- acorn_data |> 
  group_by(Tag_Num)  |> 
  filter(Date == max(Date)) |> 
  ungroup()

acorn_data |> 
  group_by(Tag_Num, Date) |> 
  mutate(row_id = row_number()) |> 
  ungroup() |> 
  complete(Tag_Num, row_id = 1:3)

df_expanded <- acorn_data_recent |> 
  group_by(Tag_Num) |> 
  mutate(row_id = row_number()) |> 
  ungroup() |> 
  complete(Tag_Num, row_id = 1:3) |> 
  group_by(Tag_Num) |> 
  fill(Date, .direction = "down") 
```

Checking number of acorns per tag

```{r}

df_expanded$Seedling_Ht_cm[is.na(df_expanded$Seedling_Ht_cm)] <- 0

df_expanded_count <- df_expanded |> 
  group_by(Tag_Num, Date) |> 
  mutate(Num_germinated = sum(Seedling_Ht_cm > 0, na.rm = TRUE)) |> 
  ungroup()

df_expanded_count |> 
  group_by(Treatment) |> 
  distinct() |> 
  count()

```

Find number of germinated acorns per tag

```{r}
#Find number of surviving acorns per tag (count tag number where seedling ht>0)
acorn_data_count <- acorn_data_recent |> 
  group_by(Tag_Num, Date) |> 
  mutate(Num_germinated = sum(Seedling_Ht_cm > 0, na.rm = TRUE)) |> 
  ungroup()

#Calculate germination rate per tag

acorn_data_count<- acorn_data_count |> 
  group_by(Tag_Num) |> 
  mutate(Germination_rate= Num_germinated/3)

acorn_data_count$Total_planted<- 3
```

Calculate germination rate per treatment

```{r}
ggplot(acorn_data_count) +
  aes(x = Treatment, y= Germination_rate, fill= Treatment) + 
  labs(x = "Treatment", y= "Germination rate") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
  geom_violin(color= "#071A0B", outlier.colour = "red")
```

**Summarize Germination rate per treatment**

```{r}

tag_germination<- acorn_data_count |> 
  select(Date, Tag_Num, Treatment, Num_germinated, Total_planted) |> 
  distinct()


#summarize total germinated, planted and germination rate
germination_count<- tag_germination |> 
  group_by(Treatment) |> 
  summarize(n=n(), total_germ_treat= sum(Num_germinated), total_planted_treat= sum(Total_planted), germination_rate= total_germ_treat/total_planted_treat*100, se_germ = sqrt((germination_rate/100) * (1 - (germination_rate/100)) / n))

germination_count<- germination_count |> 
  mutate(germination_dec= germination_rate/100)

  

```

Bar plot of germination rate per treatment + error bars

```{r}
#bar plot
ggplot(germination_count) +
  aes(x = Treatment, y= germination_dec, fill= Treatment) + 
  labs(title = "Acorn Germination Rate by Treatment", 
x = "Treatment", y= "Germination Rate (%)") +
  scale_fill_manual(
    values = c("AC" = "#052913", "DL" = "#125D6E", "DT"= "#6E6E12", "OA"= "#64875C"),
    labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )+
    scale_x_discrete(labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area"))+
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.40)) +
  geom_col(color= "#071A0B")+
    geom_errorbar(aes(ymin = germination_dec - se_germ, ymax = germination_dec + se_germ), width = 0.15)+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5), panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5))

```

Plot takeaways

-   Only 3/4 original treatments are represented (Dripline was only natural recruits)

-   No equal amounts of acorns due to some being seedlings (NR and S were removed for germination rate analysis)

-   If they're planted in the same general vicinity, there may not be a vast difference in water availability or other nutrients that would help them thrive

-   The most notable difference would be amount of shade availability but all acorns seemed to have similar success (which is why we dive into seedling height)

Other plot

```{r}
#plot with dodge bars (germinated vs not)
germination_long <- germination_count |> 
  rename(Germinated= total_germ_treat,
         Planted= total_planted_treat) |> 
  mutate(Dead=Planted-Germinated) |> 
  pivot_longer(cols = c(Germinated, Planted, Dead), 
               names_to = "Status", 
               values_to = "Amount") |> 
  filter(Status!= "Planted") 
  
  germination_long$Status[germination_long$Status == "Dead"] <- "Not Germinated"


ggplot(germination_long) +
  aes(x = Treatment, y= Amount, fill= Status) + 
  labs(x = "Treatment", y= "Number of Acorns") +
  scale_fill_manual(values = c("Germinated" = "#64875C","Not Germinated" = "#593633", name = "Status"))+
  scale_x_discrete(labels = c("AC" = "Adult Canopy", "DL" = "Dripline", "DT"= "Downed Tree", "OA"= "Open Area"))+
  geom_col(color= "#071A0B", position="dodge")
```

Why are the number of acorns different per treatment?

```{r}

```

Make a summary table with treatment, \# of tags, \# of acorns planted, \# of acorns germinated

```{r}
treatment_sum<- tag_germination |> 
  group_by(Treatment) |> 
  summarize(tag_count=n(), total_germ_treat= sum(Num_germinated), total_planted_treat= sum(Total_planted), germination_rate= total_germ_treat/total_planted_treat*100)

#make into a nice table
knitr::kable(treatment_sum)

#new table
Treatment_germination_summary<- treatment_sum
colnames(Treatment_germination_summary) [colnames(Treatment_germination_summary) == "tag_count"] <- "Number of Tags"
colnames(Treatment_germination_summary) [colnames(Treatment_germination_summary) == "total_germ_treat"]<- "Number of Germinated Acorns"
colnames(Treatment_germination_summary)[colnames(Treatment_germination_summary) == "total_planted_treat"]<- "Number of Planted Acorns"
colnames(Treatment_germination_summary)[colnames(Treatment_germination_summary) == "germination_rate"]<- "Germination Rate"


 knitr:: kable(Treatment_germination_summary)
```

Germination rate across time

```{r}
acorn_data

#number of acorns germinated per trial
acorn_data <- acorn_data |> 
  select(-Month) |> 
  mutate(month_name = month(Date, label = TRUE))

acorn_data$month_name[acorn_data$month_name == "Jan"] <- "Feb"

#summarize germination rate per treatment per month
#convert all dead to 0
acorn_data_count2 <- acorn_data |> 
  group_by(Tag_Num, month_name) |> 
  mutate(Num_germinated = sum(Seedling_Ht_cm > 0, na.rm = TRUE)) |> 
  ungroup()

acorn_data_count2$Total_planted<- 3

tag_germination2<- acorn_data_count2 |> 
  select(month_name, Tag_Num, Treatment, Num_germinated, Total_planted) |> 
  distinct()

germination_count2<- tag_germination2 |> 
  group_by(Treatment, month_name) |> 
  summarize(total_germ_treat= sum(Num_germinated), total_planted_treat= sum(Total_planted), germination_rate= total_germ_treat/total_planted_treat*100)

```

Check how many tags per month

```{r}
acorn_data_count2 |> 
  select(month_name, Tag_Num) |> 
  distinct() |> 
  group_by(month_name) |> 
  count()
```

Graph month and treatment

```{r}
month_order <- c("Jan", "Feb", "Mar", "Apr", "May")

germination_count2$month_name <- factor(germination_count2$month_name, levels = month_order)

ggplot(germination_count2, aes(x = month_name, y = germination_rate, group = Treatment, color=Treatment)) +
  geom_line() +
  geom_point()+
  labs(x = "Month", y="Survival Rate (%)")+
  scale_color_manual(
    values = c("AC" = "#06421C", "DT"= "lightblue", "OA"= "#7FB58F"),
    labels = c("AC" = "Adult Canopy", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )
```

Try

```{r}
#ggplot(germination_count2, aes(x = total_planted_treat, y = total_germ_treat, group = Treatment, color=Treatment)) +
  geom_line() +
  geom_point()+
  labs(x = "Total Planted", y="Total Germinated")+
  scale_color_manual(
    values = c("AC" = "#06421C", "DT"= "lightblue", "OA"= "#7FB58F"),
    labels = c("AC" = "Adult Canopy", "DT"= "Downed Tree", "OA"= "Open Area" ),
    name = "Treatment"
  )
```

### GLIM

Binomial distribution – 1 acorn per row, germ vs not

```{r}
head(acorn_data_count)

#make 3 rows per tag, those w seedling height of 0 --> dead
```
